---
title: "FLRef User Guide: Reference Point estimation and visualization in FLR"
author: 
 - Hennig Winker (JRC-European Commission)
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  pdf_document:
    number_sections: yes
    toc: true
    keep_tex: true
vignette: >
  %\VignetteIndexEntry{ss3daigs Vignetted}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r wrap-hook, echo = FALSE}
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
```


```{r, echo = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "  " ,fig.align = 'center', cache=FALSE,tidy.opts=list(width.cutoff=71), tidy=TRUE)

```



# Getting started {#s1}

This vignette introduces the `FLRef` R package available on https://github.com/Henning-Winker/FLRef, as a support tool for estimating and visualizing reference points in `FLR`. Specific emphasis is put to enable routine plotting of a wider a of biological reference points (BRPs), such, as $F_{spr40}$, $F_{B35}$ or $F_{0.1}$.  

## Installation

`FLRef` requires very recent versions of `FLR` libraries `FLCore`, `FLBRP`, `FLasher`, `mse`, `FLSRTBM` and `ggplotFL`. This can be installed together with `FLRef` from gihtub using library(devtools):

```{r, eval=FALSE}
installed.packages("devtools")

installed.packages("ggplot2")

installed.packages("ggpubr")

installed.packages("TMB")


devtools::install_github("flr/FLCore")

devtools::install_github("flr/FLBRP")

devtools::install_github("flr/FLasher")

devtools::install_github("flr/mse")

devtools::install_github("flr/ggplotFL")

devtools::install_github("flr/FLSRTMB")

devtools::install_github("henning-winker/FLRef")

# only for demo
install.packages("ggpubr")


```


```{r,message=FALSE}
library(FLCore)
library(FLBRP)
library(FLasher)
library(FLSRTMB)
library(ggplotFL)
library(FLRef)
library(ggpubr) # For this demo
```


<br>

## Example stock

The North Sea Plaice `FLStock` object `ple4` from `FLCore` used here as an example. 

```{r, eval=TRUE}

data(ple4)
stk = ple4
```
```{r fig1, fig.height=4, fig.cap = "Estimated stock trajectories"}
plot(stk)+theme_bw()+ facet_wrap(~qname,scales="free",ncol=2)
```

<br>
\newpage

# Per-recruit reference points

Common proxies for $F_{MSY}$ that do not neccessarly require a stock recruitment relationship are $F_{0.1}$ and $F_{SPR35-50}$, where $SPR$ the spawning ratio potential expressed as spawning-biomass per-recruit relative to the unfished spawning biomass per-recruit at $F=0$ ($SPR_0$). $F_{SPR40}$ denotes a spawning-biomass per-recruit is reduced to 40 percent of $SPR_0$.

A range of these $F_{BRP}$'s can be computed quickly by:


```{r, eval=TRUE}

fbrps = computeFbrps(stock=ple4, proxy="sprx",f0.1=TRUE, verbose = FALSE)

```

<br>

This range $F_{BRP}$ values can easily visualised


```{r fig2, fig.height=4.5, fig.cap = "Estimated per-recruit reference points corresponding to the $F_{BPR}$'s $F_{spr35-50}$ and $F_{0.1}$."}

ploteq(fbrps)

```
 
<br>

$Yield$ and $SSB$ are in this case as yield- and spawning biomass per-recruit, respectively. $B0$ is the product of $R_0$ and $SPR_0$, where $SPR_0$ is a function of weight-at-age ($w_a$), maturity-at-age ($mat_a$) and natural mortality at age ($M_a$). Because $R_0$ is one (per-recruit), $B0$ equals $SPR_0$.   
It is also possible to add some of the "default" reference points that are inbuilt in `FLBRP`. 
 
```{r fig3, fig.height=4.5, fig.cap = "Estimated per-recruit reference points corresponding to the $F_{BPR}$, $F_{spr35-50}$ and $F_{0.1}$, adding the inbuilt default reference point $F_{max}$."}

ploteq(fbrps,refpts = "fmax")

``` 
 
A more targeted approach for exploring option of target an limit reference points is the function `computeFbrp()` (i.e. without 's). In the following example the $F_{brp}$ is chosen to be $F_{0.1}$ and a $B_{lim}$ proxy is chose so that is corresponds $0.25B_{F0.1}$.


```{r, eval=TRUE}

fbrp = computeFbrp(stock=stk, proxy=c("f0.1"),blim=0.25,type="btrg", verbose = FALSE)

Fbrp(fbrp)

```

It is also possible to add additional $F_{BRP}$. However, note that by convention the first in order of occurrence is used, e.g. to compute the ratio to approximate $B_{lim}$. It the example below $B_{lim}$ is now computed as $0.25B_{spr40}$, i.e relative to the biomass per-recruit corresponding to $F_{spr40}$, spefied as `proxy = "sprx"` and `x=40`.   


```{r, eval=TRUE}

fbrp = computeFbrp(stock=stk, proxy=c("sprx","f0.1"),x=40,blim=0.25,type="btrg", verbose = FALSE)

Fbrp(fbrp)

```
 
```{r fig4, fig.height=4.5, fig.cap = "Estimated per-recruit reference points corresponding to the"}

ploteq(fbrp,refpts = "fmax")

``` 
 
<br>

The `plotAdvice()` function provide can then be used to show the estimated stock trajectories per recruit relative to the reference points. To compute those from the `FLStock` object, the recruitment is normalised by its geometric mean which is a assumed to approximate $R_0$ (i.e. expected mean recruitment in the absence of a stock recruitment relationship). The estimate of spawning biomass per recruit is computed as $SB/R=SSB/R_{0}$ and then expressed as the Spawning Ratio Potential (SRP) relative to $SPR_0$. The "observed" yield per recruit is first computed as $Y/R=landings/R_{0}$ and then expressed as the ratio to the equilibrium Yield corresponding to $F_{BRP}$.    


```{r fig5, fig.height=4.5, fig.cap = "Stock advice plot showing the modelled quantities from a per-recruit perspestive relative to per-recruit based reference points"}

plotAdvice(stk,fbrp)

``` 

<br>

\pagebreak

# Integrating stock recruitment (S-R) functions into reference point computations

The simplest S-R model is assuming a that the expected recruitment is constant with $R_0$ estimated in the form of the geometric mean. This `geomean` can therefore be interpreted as Null model S-R functions. To set this up in `FLSRTMB`, it is only required to create a standard `FLSR` object as input to the function `srrTMB()`:

```{r, eval=TRUE}

object = as.FLSR(stk,model=geomean)

gm = srrTMB(object)

```

<br> 

The reference points can now be re-calculated with `computeFbrp()` by simply specificying `sr=gm`, such that



```{r, eval=TRUE}

fbrp = computeFbrp(stock=stk,sr=gm, proxy=c("sprx","f0.1"),x=40,blim=0.25,type="btrg", verbose = FALSE)
```

The only difference to the per-recruit representation is that that the reference points to recruitment, biomass and yield are now readily scaled by $R_0$ to the corresponding modelled quantities, which allows to add those for reference using the option `obs=TRUE`.     

```{r fig6, fig.height=4.5, fig.cap = "Estimated  reference points relative to estimates of $Recruitment$, $SSB$, $F$ and $Landings$"}

ploteq(fbrp,refpts="fmax",obs=TRUE)

```
<br>

Similarly, the estimated time-series of $Recruitment$, $SSB$, $F$ and $Landings$ can now be directly compared to the reference points on absolute scale. Otherwise, the inference about the stock status remains the same as for the per-recruit analysis in the absensce a S-R relationship. 

<br>

```{r fig7, fig.height=4.5, fig.cap = "Stock advice plot showing modelled quantities and the corresponding reference points"}

plotAdvice(stk,fbrp)

``` 

<br>

\newpage

\pagebreak

The next step is to set fit alternative S-R functions with `srrTBM()`

The first one is a `model=bevholtSV` which is parameterised as a function of steepness $s$ and $SPR_0$. This formulation also requires to specify `spr0 = spr0y(stk)`, which computes the implicit values of $SPR_0$ in each year $y$ as function of $w_{a,y}$, $mat_{a,y}$ and $M_{a,y}$. The estimates of $s$ and $R_0$ are subsequently converted into the conventional `bevholt` parameter `a` and `b` given the mean $SPR_0$ for some reference years. For example, the default is use the geometric mean $SPR_0$ over the time-series whereas specifying `nyears=3` would use the mean of $SPR_0$ over the 3 most recent years.       

```{r, eval=TRUE}

bh = srrTMB(as.FLSR(stk,model=bevholtSV),spr0=spr0y(stk),verbose = FALSE)

bh@SV
```

<br>

Calling bh@SV shows the maximum likelihood estimates of $s$, the recruitment standard deviation $sigmaR$, $R0$ and the post-hoc computed AR1 auto-correlation coeffecient $rho$.

Similarly, the Ricker model `model=rickerSV` is parameterised as a function of steepness $s$ and $SPR_0$, but $s$ is in this case not restricted to an bound at one to enable obtaining the same unconstraint fits as the equivalent $a$, $b$ formulation of the model. 


```{r, eval=TRUE}

ri = srrTMB(as.FLSR(stk,model=rickerSV),spr0=spr0y(stk),verbose = FALSE)

```

Finally, `FLSRTMB` also allows to fit a hockey-stick `model=segreg`, which is formulated as function of $SPR_0$. This formulation enables to invoke contraits for the location of the break-point. For example, by specifying `lplim=0.05` and `uplim=0.2` the location of the break-point $b=B_{lim}$ is constrained to fall between $0.05-0.2B_0$, which is in the specific case of the hockey-stick identical to the spawning ratio potential $SRP_{0.05-0.2}$. 

```{r, eval=TRUE}

hs = srrTMB(as.FLSR(stk,model=segreg),spr0=spr0y(stk),lplim=0.05,uplim=0.2)

```

The three S-R fit can be summarised in single $FLSRs$ to enable a quick comparison with `plotsrs`.

```{r fig8, fig.height=3.5, fig.cap = "Comparison of the S-R relationship fitted by assuming Beverton-Holt (bh), Ricker (ri) and Hockey-Stick (hs) function. Open circles show the observed S-R pairs with the solid do denoting the final assessment year"}

srs = FLSRs(bh=bh,ri=ri,hs=hs)

plotsrs(srs)

```

<br> 
\newpage

Clearly, the hockey-stick fails to identify a clear break point in the data and therefore is located towards the lower specified bound, `lplim=0.05`.


```{r, eval=TRUE}

hs@SV[["BlimB0"]]

```
The stock shows a considerable variation and by providing a vector of `spr0=spr0y(stk)` the model effectly assumes time-varying $SPR_{0_y}$ and thus $B_{0_y}$. 

```{r fig8a, fig.height=3.5, fig.cap = "Annual $SPR_{0_y}$ as function of time-varying $w_{a_y}$ (here), $mat_{a_y}$ and $M_{a_y}$.",tidy=FALSE}

plot(spr0y(stk))+theme_bw()+
  ylab(expression(SPR[0]))+xlab("Year")+
  geom_hline(yintercept = mean(spr0y(stk)),linetype="dashed")

```

An alternative is to set $SPR_0$ to its mean or change the bounds $lplim$ and $uplim$, which determine the "plausible" range of $SRP$. For this example, the lower limit of $lplim$ is increase to 0.07.  

```{r fig8b, fig.height=3.5, fig.cap = "Comparison of the Hockey-Stick specified with (1) $SRP_{5-20}$ and time-varying $SPR_{0,y}$, (2) the same but with the mean of $SPR_{0,y}$ (3) $SRP_{7-20}$ and mean $SPR_{0,y}$."}


hs1 = srs$hs
hs2 = srrTMB(as.FLSR(stk,model=segreg),
          spr0=mean(spr0y(stk)),lplim=0.05,uplim=0.2)
hs3 = srrTMB(as.FLSR(stk,model=segreg),
             spr0=mean(spr0y(stk)),lplim=0.07,uplim=0.2)

plotsrs(FLSRs(plim0.05=hs1,muSPR0=hs2,plim0.07=hs3))

```

Here models (1) `plim0.05` (2) `muSPR0` produce the same results. In option (3) the break-point is still located close to $plim = 0.07$. Therefore, the data hold no information about a break-point and the choice of "plausible" $SPR_{0}$ specification (mean vs time-varying) and the $SRP$ bounds determine the estimate of the break-point. For this demo, option (2) is used instead of (1) for subsequent illustrations. 

```{r, eval=TRUE}

hsblim(hs1)
hsblim(hs2)
hsblim(hs3)


```

```{r, eval=TRUE}

# Extract Blim
blim = c(params(hs)["b"])
# check break-point relative to B0
hsblim(hs)["SRPlim"]
hsblim(hs3)["SRPlim"]

```

The function `plotsrs` provides following options to illustrate the S-R:

+ no S-R observations `path=FALSE`

+ with S-R observations `path=TRUE`

+ Projected through to $B_0=R_0SPR_0$ 

+ Relative to $SSB_0$ and $R_0$ (permist comparson accross stocks)


```{r fig9, fig.height=5.5, fig.cap = "Comparison of the S-R relationship fitted by assuming Beverton-Holt (bh), Ricker (ri) and Hockey-Stick (hs) function."}

p1 = plotsrs(srs,path=FALSE)
p2 = plotsrs(srs,path=TRUE)
p3 = plotsrs(srs,b0=TRUE)
p4 = plotsrs(srs,b0=TRUE,rel=TRUE)

ggarrange(p1, p2, p3, p4, ncol=2, nrow=2, common.legend = TRUE, legend="right")

```

<br>

Similar to `FLSRs`,the `computeFbrp` output in the form  `FLBRP` objects can also be compiled in `FLBRS` to enable comparison. Note that in the case of the hockey-stick its breakpoint is used directly as input of an absolute value for blim, using the option `type="value"`.

```{r fig10, fig.height=5., fig.cap = "Estimated reference points at equilibrium $Recruitment$, $SSB$, $F$ and $Landings$"}
brps = FLBRPs(
  bh = computeFbrp(stk,sr=srs[["bh"]],proxy=c("f0.1","sprx","msy"),x=40,blim=0.25,type="btrg",verbose = FALSE),
  ri = computeFbrp(stk,sr=srs[["ri"]],proxy=c("f0.1","sprx","msy"),x=40,blim=0.25,type="btrg",verbose = FALSE),
  hs = computeFbrp(stk,sr=srs[["hs"]],proxy=c("f0.1","sprx","msy"),x=40,blim=blim,type="value",verbose = FALSE)
)

# plot
ploteq(brps)

```


\newpage

<br>

The same plot can be produce with estimates from the assessment estimates.


```{r fig10a, fig.height=5., fig.cap = "Estimated  reference points relative to estimates of $Recruitment$, $SSB$, $F$ and $Landings$"}

# plot
ploteq(brps,obs=TRUE)

```


<br>

`FLSRTMB` provides also the option fix $s$ or use informative priors, such as those that can be derived from [`FishLife`; Thorson (2020)](https://github.com/James-Thorson-NOAA/FishLife). This can be done 

```{r fig11, fig.height=3.5, fig.cap = "Comparison of alternative parameterisation of the Beverton Holt S-R"}

# Fixed steepness
s = c(seq(0.8,0.95,0.05))
fixs = FLSRs(lapply(as.list(s),function(x){
  srrTMB(as.FLSR(stk,model=bevholtSV),spr0=spr0y(stk),s=x,s.est = FALSE)
}))
names(fixs) = paste0("s=",s)
# with prior with mean s=0.85 and s.logitsd = 0.3
s.pr = srrTMB(as.FLSR(stk,model=bevholtSV),spr0=spr0y(stk),s=0.8,s.logitsd = 0.3)
# uncontrained estimate
s.est = srrTMB(as.FLSR(stk,model=bevholtSV),spr0=spr0y(stk),s=0.8)

#combine
bhs = FLSRs(c(s.est=s.est,s.pr=s.pr,fixs))
# add s estimate
names(bhs)[1:2] = c(paste0("s.est(",round(s.est@SV[["s"]],2),")"),
  paste0("s.pr(",round(s.pr@SV[["s"]],2),")"))

plotsrs(bhs)

```


```{r fig12, fig.height=4.5, fig.cap = "Comparison of equilibriun curves and reference points for alternative parameterisation of the Beverton Holt S-R"}

bh.brps = FLBRPs(lapply(bhs,function(x){
  computeFbrp(stk,x,proxy=c("f0.1","msy"),blim=0.25,type="btrg",verbose=FALSE)
}))

ploteq(bh.brps,obs=TRUE,panels=4)

```

```{r fig13, fig.height=4.5, fig.cap = "Stock advice plot showing modelled quantities and the corresponding reference points for a Beverton S-R model with estimated $s$"}

plotAdvice(stk,bh.brps[[1]])+
  ggtitle(paste0(stk@name,": BevHolt with s = ",round(bhs[[1]]@SV[["s"]],3)))

```

Another option to illustrate the stock status against the reference point estimates is the "Advice Rule" plot `plotAR()`. For the variety option please see the available examples `?plotAR`. Here we consider 4 options of illustration: (1) Basic plot with a precautionary biomass $B_{pa}$ add that expressed relative $B_(lim)$,
adding a $B_{trigger}$ as fraction of the target Biomass reference point $Btrg$, (3) using kobe type color-coding with de facto fishing closure at $B{lim}$ and (4) showing the quatative relative to the targer reference points. The input can be either the output of `Fbrp()` (easy to manipulate) or the `FLBRP` output from `computeFbrp()`. 


```{r fig14, fig.height=4.5, fig.cap = "Stock advice plot showing modelled quantities and the corresponding reference points for a Beverton S-R model with estimated $s$"}

pars= Fbrp(bh.brps[[1]])
pars
p1= plotAR(bh.brps[[1]],obs=stk,kobe=FALSE,bpa=1.4)
p2= plotAR(bh.brps[[1]],obs=stk,kobe=FALSE,
           bpa=1.4,btrigger=0.7)
p3= plotAR(bh.brps[[1]],obs=stk,kobe=TRUE,
           bpa=1.4,btrigger=0.7,bclose=1,fmin=0.01)
p4= plotAR(bh.brps[[1]],obs=stk,kobe=TRUE,
           bpa=1.4,btrigger=0.7,rel=TRUE)

ggarrange(p1, p2, p3, p4, ncol=2, nrow=2)

```

